Investigating the Molecular Processes Behind the Cell-Specific Toxicity Response to Titanium Dioxide Nanobelts

### 2021-08-29 manuscript.tex
Version Laurent and I think is ready for resubmitting, allowing us to check this commit.

% Authors, for the paper (add full first names)
\Author{\hl{Laurent A. Winckers} %MDPI: Please carefully check the accuracy of names and affiliations.
 $^{1}$\orcidA{}, Chris \hl{T.} %MDPI: we added dot for middle name abbreviation, please confirm.
 Evelo $^{1,2}$\orcidB{}, Egon \hl{L.} Willighagen $^{1}$\orcidC{} and Martina Kutmon $^{1,2,}$*\orcidD{}}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Department of Bioinformatics\hl{---}%MDPI: changed from hyphen into em dash.
BiGCaT, NUTRIM School of Nutrition and Translational Research in Metabolism, Maastricht \hl{University}; %MDPI: PLEASE add city, state abbreviation postcode, country.
 \hl{laurent.winckers@maastrichtuniversity.nl (L.A.W.); chris.evelo@maastrichtuniversity.nl (C.T.E.);  egon.willighagen@maastrichtuniversity.nl (E.L.W.)} %MDPI: Newly added information, please confirm.
\\
$^{2}$ \quad Maastricht Centre for Systems Biology (MaCSBio), Maastricht \hl{University}}

@@

The detailed biological processes related to the toxicity of many engineered nanomaterials are not yet all fully understood~\cite{Podila2012,Bahadar2016}. Due to the varied physicochemical properties of nanomaterials and often even of the particles within the nanomaterial itself, it is difficult to predict the biological effects leading to toxicity. Biological response and toxicity depend on the size of the nanoparticle, size of the agglomerate, surface impurities, and~degradability~\cite{DeStefano2012}. Furthermore, other factors that have an effect on biological response and toxicity include the method of exposure, the~entry route into the human body, and~the distribution in the body ~\cite{Shin2015}. %check meaning has been retained - duplicated text removed as well as minor edits to this sentence
The~shape of a nanoparticle also has an influence on the nanoparticle’s effects, where for example nanobelts, nanostructures in the form of belts, have been shown to be more pro-inflammatory compared to spheres~\cite{Silva2013}. Studies have found that there are various biological effects and toxicities of engineered nanomaterials under different circumstances and with varying engineered nanomaterials~\cite{Ray2009,Atha2017,Donaldson2003,Sung2008,EbabeElle2013}. These changes in biological effects are ascribed to the differences in chemical and physical properties of nanomaterials. However, when studying the properties, biological effect relations are complicated because of the difficulty in creating identical nanomaterials from different batches and/or sources due to the varying physicochemical properties of nanomaterials~\cite{Mulhopt2018}. %check meaning has been retained
Nevertheless, manual curation of transcriptomics datasets regarding nanomaterials is performed to enhance their degree of 
fairness  %check meaning has been retained
to aid nanomaterial safety assessment~\cite{AliisaSaarimaki}.

@@

% start a new page without indent 4.6cm
%\clearpage
\end{paracol}
\nointerlineskip
\begin{figure}[H]
\widefigure
\includegraphics[width=0.83\linewidth]{fig1.png}
  \caption{\hl{Gene} %MDPI: please modify hyphen into minus in the image, same as in other figures.
 expression volcano plots for different {cellular models} and TiO\textsubscript{2}-nanobelt concentrations.
   On the x-axis log\textsubscript{2} (fold change) is depicted whereas on the y-axis the \hl{-}%MDPI: please confrim if this should be minus sigh $-$.
log\textsubscript{10} (\emph{p}-value) is depicted. The~dotted lines represent cut-off values for significantly changed genes (log\textsubscript{2} fold change $>$ 0.26 or $<$ $-$0.26, \emph{p}-value $<$ 0.05). A brown color depicts that a gene meets both cut-off criteria, a~blue color relates to meeting only the \emph{p}-value cut-off, an~orange color relates to meeting only the log$_{2}$ fold change cut-off, and a gray color indicates that a gene does not meet any of the criteria.}
\label{fig:fig1}
\end{figure}
\begin{paracol}{2}
%\linenumbers
\switchcolumn
\vspace{-12pt}

@@

\subsection{Toxicity-Related Pathways}
To gain more insight into the toxicity-related processes, the~overrepresented pathways were further categorized into pathways related to the apoptotic process, inflammatory response, DNA damage response, and/or oxidative stress. Often this kind of clustering of the pathways is performed manually. However, we implemented a gene-based approach. First, gene sets of four Gene~Ontology~(GO) terms were obtained, i.e.,~“apoptotic process”~(GO\hl{:}%MDPI: please confirm if space should be added after ``:''.
0006915, 1269 genes), “inflammatory response”~(GO\hl{:}0006954, 569 genes),  “cellular response to DNA damage stimulus”~(GO\hl{:}0006974, 762 genes), and~“response to oxidative stress”~(GO\hl{:}0006979, 243 genes). Evidently, these processes are tightly connected. Whereas some can be causative for others, they are expected to overlap. Additional Figure~\ref{fig:figA1} shows the gene overlap between the four GO gene sets in a Venn diagram. Next, we calculated the gene overlap between the pathways from the WikiPathways pathway collection with the annotated genes of the four GO-terms. The~overlap was calculated by dividing the number of toxicity-related genes present in the pathway by the total number of genes present in the pathway. As~a cut-off we used 50$\%$ indicating that at least half of the pathway is directly linked to the GO-term of interest via their gene~overlap.

To select the desired gene overlap cut-off, we compared four cut-offs with each other and an overrepresentation analysis approach to see how many pathways these would give. The~cut-offs we used were 50\%, 60\%, 70\%, and 80\%. For~the overrepresentation analysis we looked for overrepresentation of the pathway genes in the gene lists of the GO-terms. The~results are shown in additional Table~\ref{tab:tabA1}. The~overrepresentation analysis \mbox{(\emph{p}-value~\hl{< $<$}~0.05)} %Please confirm if this is correct.
showed the most pathways to be linked to the four GO-terms; in~this case it showed 401, 250, 203, and 205 pathways overrepresented for the apoptotic process, inflammatory response, DNA damage, and oxidative stress GO-terms, respectively. The~80\% and 70\% gene overlap resulted in almost similar amounts of pathways compared to each other. They even resulted in 0 pathways for the inflammatory response GO-term. Moreover, it stands out that for the gene overlap approach only one pathway meets at least 50\% or higher gene overlap with the oxidative stress GO-term. Additionally, it is interesting to see that an overrepresentation analysis prompts hundreds of pathways compared to just dozens when looking at gene overlap. Comparing the results of the overrepresentation analysis and various cut-offs, the at least 50\% gene overlap cut-off was deemed best for our~approach.

After this filtering step we found that, out of the 1076 pathways in the human pathway collection, 66 pathways are related to the apoptotic process, 15 are linked to inflammatory response, 28 are linked to DNA damage, and 2 are linked to oxidative stress.
It is noticeable that there are only two oxidative stress pathways, i.e.,~Detoxification of Reactive Oxygen Species~(wikipathways\hl{: } %MDPI: added space, please confirm.
WP2824~\cite{WP2824}) and Oxidative Stress~(wikipathways: WP408~\cite{WP408}), that have at least 50$\%$ of their genes overlapping with the genes annotated to the GO-term ``response to oxidative stress''. Based on this finding we can argue that most pathways in the pathway database we have used describe different processes that result in oxidative stress or where oxidative stress is part of/has an influence on the process itself. However, these do not describe oxidative stress as a process.  Nevertheless, these pathways are of interest for our~analysis.

@@

\subsection{{Caco-2/HT29-MTX} Cells}
While for both concentrations the {Caco-2/HT29-MTX co-culture} shows the same number of overrepresented pathways, the~pathways compared between the low and high concentrations are not all the same. Of~the ten pathways, three are found for both concentrations, namely IL-5 Signaling Pathway (wikipathways: WP127,~\cite{WP127}), IL-2 Signaling Pathway (wikipathways: WP49,~\cite{WP49}) and Endometrial Cancer (wikipathways: WP4155,~\cite{WP4155}). 
Interestingly, the low concentration pathways related to DNA-damage and repair show up in the results. %check meaning has been retained
For~example, HDR through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways: WP3567,~\cite{WP3567}), Nonhomologous End-Joining (NHEJ) (wikipathways: WP3550,~\cite{WP3550}), and DNA Double-Strand Break Response (wikipathways: WP3543,~\cite{WP3543}) are among these pathways. However, for the high concentration it is interesting to see that the Apoptosis pathway (wikipathways: WP254,~\cite{WP254}) shows up, which could be an indication of {Caco-2/HT29-MTX} cells possibly undergoing apoptotic processes upon exposure to TiO\textsubscript{2}-nanobelts. However, these results indicate that the studied processes in {Caco-2/HT29-MTX co-cultures} are not extensively affected by the administration of TiO\textsubscript{2}-nanobelts, for~either the low or the high concentration. This finding supports the conclusion based on the cell viability assay results in the original study, which showed that there was no significant decrease in cell viability for the {Caco-2/HT29-MTX co-culture} for both concentrations of TiO\textsubscript{2}-nanobelts~\cite{Tilton2013}.

\begin{figure}[H]
\includegraphics[width=1\linewidth]{fig2.pdf}
  \caption{Pathway-gene networks of altered toxicity-related pathways.
   The color of the nodes indicates to which GO-term the pathway is affiliated. Orange indicates apoptosis, blue indicates inflammation, pink indicates DNA damage, and 
   red %check meaning has been retained
   indicates oxidative stress. Gray nodes represent genes. Number (n) represents the number of significantly overrepresented pathways that are depicted in the networks.}
\label{fig:fig2}
\end{figure}
\unskip
\subsection{SAE Cells}
For the SAE cell line, the~low concentration only prompted three overrepresented pathways compared to 15 for the high concentration. Noticeable as well, only one of those three pathways was found to be overrepresented also for the high concentration, namely HDR through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways: WP3567,~\cite{WP3567}). This increase in pathways with a significantly increased overrepresentation of affected genes is an indication that upon exposure to the high concentration of TiO\textsubscript{2}-nanobelts, more biological processes related to toxicity are affected compared to the low~concentration.

Next, for~the SAE cell line exposed with the low concentration, we found three overrepresented pathways. As~for the high concentration we found 15 overrepresented pathways. This increase in pathways with a significantly increased overrepresentation of affected genes is an indication that more biological processes related to toxicity are affected compared to the low concentration upon exposure to the high concentration of TiO\textsubscript{2}-nanobelts . For~the low concentration the pathways HDR through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways: WP3567,~\cite{WP3567}), IL-6 signaling pathway (wikipathways: WP364,~\cite{WP364}), and Regulation of TP53 Activity through Phosphorylation (wikipathways: WP3838,~\cite{WP3838}) are overrepresented. The~first pathway describes a process of DNA double-strand break repair, where the second pathway describes the signaling of the cytokine IL-6 and the last pathway describes the regulation of TP53. Except~for the first pathway, which is also found in the result for the high concentration, the~overrepresented pathways describe more general regulatory pathways rather than detailed process-describing pathways. However, for~the high concentration we see multiple pathways related to DNA damage, such as the aforementioned HDR through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways: WP3567,~\cite{WP3567}), DNA IR-damage and cellular response via ATR (wikipathways: WP4016,~\cite{WP4016}), DNA IR-Double Strand Breaks (DSBs) and cellular response via ATM (wikipathways: WP3959,~\cite{WP3959}), and DNA Mismatch Repair (wikipathways: WP531,~\cite{WP531}) among others. Moreover, the results prompted apoptosis-related pathways such as Apoptosis Modulation and Signaling (wikipathways: WP1772,~\cite{WP1772}) and Apoptosis Modulation by HSP70 (wikipathways: WP384,~\cite{WP384}). This is also an indication that the high concentration induced more distinct biological processes related to toxicity compared to the low concentration for the SAE cell~line.

\subsection{THP1 Cells}
For the THP1 cell line, there are only 10 significantly overrepresented pathways for the low concentration of TiO\textsubscript{2}-nanobelts. However, among~these pathways, the~pathways TNF signaling (wikipathways: WP3380,~\cite{WP3380}), Apoptosis (wikipathways: WP254,~\cite{WP254}), Apoptotic execution phase (wikipathways: WP1784,~\cite{WP1784}), TNF alpha Signaling Pathway (wikipathways: WP231,~\cite{WP231}), and TNF-related weak inducer of apoptosis (TWEAK) Signaling Pathway (wikipathways: WP2036,~\cite{WP2036}) are present. This indicates that upon exposure to the low concentration of TiO\textsubscript{2}-nanobelts the THP1 cell line activates immune- and inflammation-related processes. This result can be explained as a general cell activity response since THP1 is a macrophage-like cell line. However, it can also be explained due to the effect of TiO\textsubscript{2}-nanobelts on the THP1 cell line in this case. TiO\textsubscript{2}-nanobelts are likely to induce toxic processes, even at a low concentration, which results in an inflammatory response of the THP1 cells. Similar inflammatory responses, like activation of Nf-$\kappa$B and production of TNF-$\alpha$, were seen in these cells upon exposure to other nanoparticles such as ZnO NM-110, SiO\textsubscript{2} NM-200, and Ag NM-300~\cite{Brzicova2019}.

Compared to the low concentration, the~high concentration yields more pathways with significant overrepresentation i.e.,~39 versus 10. Except~for the pathway Apoptotic execution phase (wikipathways: WP1784,~\cite{WP1784}) all other 9 pathways which showed up for the low concentration were found for the high concentration as well. In~addition to these pathways, pathways such as Oxidative Stress (wikipathways: WP408,~\cite{WP408}), Nanoparticle triggered regulated necrosis (wikipathways: WP2513,~\cite{WP2513}), and Mismatch Repair (wikipathways: WP3381,~\cite{WP3381}) %please be consistent with capitalization of pathways throughout the paper (at present some have initial capitals and some do not)
and more were among the significant pathways. These results indicate that the THP1 cell line upon exposure to the high concentration of TiO\textsubscript{2}-nanobelts results in toxicity-related processes such as inflammation, DNA damage, and~oxidative stress. The~higher number of pathways that show up in the result together with the biological processes they depict indicate that the high concentration induces greater effects than the low concentration. This was also seen in the original paper where the low concentration induced a significant decrease in cell viability, while the high concentration induced an even greater decrease~\cite{Tilton2013}. Furthermore, it has also been shown that nanoparticles can activate similar processes, such as inflammatory processes and DNA damage, as~seen upon exposure to the high concentration~\cite{Brzicova2019,Huang2017}.

{The THP1 cell line is the cellular model that has the most inflammation-related pathways in the results: four for the low concentration and nine for the high concentration. All four of the low concentrations show up in the results of the high concentration as well. These pathways are Photodynamic-therapy-induced NF-kB survival signaling} (wikipathways: WP3617,~\cite{WP3617}), {Interleukin-10 signaling} (wikipathways: WP4063,~\cite{WP4063}), {Fibrin Complement Receptor 3 Signaling Pathway} (wikipathways: WP4136,~\cite{WP4136}), {and Platelet-mediated interactions with vascular and circulating cells} (wikipathways: WP4462,~\cite{WP4462}).
{Nanoparticle-induced genotoxicity can arise and a distinction between primary and secondary genotoxicity can be made. Inflammation drives secondary toxicity as suggested by} Emma Åkerlund~et~al., {who showed that conditioned media from differentiated THP1 cells induced DNA-damage in HBEC after} 3~h~exposure~\cite{Akerlund2019}. {While there are inflammatory-related pathways in the results for the THP1 cell line, only the high concentration has three DNA-damage-related pathways in the results. The~presence of DNA-damage-related pathways and inflammatory-related pathways could indicate secondary genotoxicity for the THP1 cells upon exposure to the high concentration. However, more research is needed to address this exact mechanism.}

@@

To illustrate how the effects can be studied in more depth we visualized the log fold change of all six conditions on the Oxidative Stress pathway (wikipathways: WP408~\cite{WP408}), the~Apoptosis pathway (wikipathways: WP254~\cite{WP254}), and the DNA Mismatch Repair pathway (wikipathways: WP531~\cite{WP531}). This shows that for most genes in these pathways gene expression data are present (see additional Figures~\ref{fig:figA2} and~\ref{fig:figA3}).

@@

\begin{figure}[H]
\includegraphics[height=14cm,keepaspectratio]{fig3.png}
  \caption{\hl{Visualization} of log fold change on the Oxidative Stress pathway (wikipathways: WP408) for all six conditions.
{Cellular models} are depicted from left to right as {Caco-2/HT29-MTX}, SAE, and THP1. The~top row depicts the low concentrations whereas the bottom row depicts the high concentrations. {Color gradient for differential gene expression after exposure goes from blue (downregulated) via white (not changed) to red (upregulated).}}
\label{fig:fig3}
\end{figure}
\vspace{-12pt}

@@

\subsection{Reproducible Analysis Workflow}
The complete analysis is automated in R (version 3.6.3) and can easily be repeated with a new transcriptomics dataset or different selection focus. The~R scripts are available on GitHub (\url{https://github.com/laurent2207/TiO2-scripts} \hl{(accessed on)}) and archived on Zenodo~\cite{TiO2-scripts}.

@@

\bibitem[Bohler \em{et~al.}(2020)Bohler, {Reactome Team}, Weitz, and
  Kutmon]{WP2824}
Bohler, A.; {Reactome Team}; Weitz, E.; Kutmon, M.
\newblock Detoxification of Reactive Oxygen Species (Homo Sapiens). 2020.
\newblock  Available online: \url{https://wikipathways.org/instance/WP2824_r117697} \hl{(accessed on). } %Please add accessed date.

>>>>

% Authors, for the paper (add full first names)
\Author{\hl{Laurent A. Winckers} %MDPI: Please carefully check the accuracy of names and affiliations.
 $^{1}$\orcidA{}, Chris \hl{T.} %MDPI: we added dot for middle name abbreviation, please confirm. Correct.
 Evelo $^{1,2}$\orcidB{}, Egon \hl{L.} Willighagen $^{1}$\orcidC{} and Martina Kutmon $^{1,2,}$*\orcidD{}}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Department of Bioinformatics\hl{---}%MDPI: changed from hyphen into em dash.
BiGCaT, NUTRIM School of Nutrition and Translational Research in Metabolism, Maastricht \hl{University}, Maastricht, NL-6200 MD, The Netherlands; %MDPI: PLEASE add city, state abbreviation postcode, country.
%Reply: Added city, postcode and country.
 \hl{laurent.winckers@maastrichtuniversity.nl (L.A.W.); chris.evelo@maastrichtuniversity.nl (C.T.E.);  egon.willighagen@maastrichtuniversity.nl (E.L.W.)} %MDPI: Newly added information, please confirm.
%Reply: looks good
\\
$^{2}$ \quad Maastricht Centre for Systems Biology (MaCSBio), Maastricht \hl{University}, Maastricht, NL-6200 MD, The Netherlands}
%Reply: Added city, postcode and country.

@@

The detailed biological processes related to the toxicity of many engineered nanomaterials are not yet all fully understood~\cite{Podila2012,Bahadar2016}. Due to the varied physicochemical properties of nanomaterials and often even of the particles within the nanomaterial itself, it is difficult to predict the biological effects leading to toxicity. Biological response and toxicity depend on the size of the nanoparticle, size of the agglomerate, surface impurities, and~degradability~\cite{DeStefano2012}. Furthermore, other factors that have an effect on biological response and toxicity include the method of exposure, the~entry route into the human body, and~the distribution in the body ~\cite{Shin2015}. %check meaning has been retained - duplicated text removed as well as minor edits to this sentence. Correct.
The~shape of a nanoparticle also has an influence on the nanoparticle’s effects, where for example nanobelts, nanostructures in the form of belts, have been shown to be more pro-inflammatory compared to spheres~\cite{Silva2013}. Studies have found that there are various biological effects and toxicities of engineered nanomaterials under different circumstances and with varying engineered nanomaterials~\cite{Ray2009,Atha2017,Donaldson2003,Sung2008,EbabeElle2013}. These changes in biological effects are ascribed to the differences in chemical and physical properties of nanomaterials. However, when studying the properties, biological effect relations are complicated because of the difficulty in creating identical nanomaterials from different batches and/or sources due to the varying physicochemical properties of nanomaterials~\cite{Mulhopt2018}. %check meaning has been retained. Correct.
Nevertheless, manual curation of transcriptomics datasets regarding nanomaterials is performed to enhance their degree of 
fairness  %check meaning has been retained. Correct.
to aid nanomaterial safety assessment~\cite{AliisaSaarimaki}.

@@

% start a new page without indent 4.6cm
%\clearpage
\end{paracol}
\nointerlineskip
\begin{figure}[H]
\widefigure
\includegraphics[width=0.83\linewidth]{fig1.png}
  \caption{\hl{Gene} %MDPI: please modify hyphen into minus in the image, same as in other figures.
% Reply: ???
 expression volcano plots for different {cellular models} and TiO\textsubscript{2}-nanobelt concentrations.
   On the x-axis log\textsubscript{2} (fold change) is depicted whereas on the y-axis the \hl{-}%MDPI: please confrim if this should be minus sigh $-$.
%Reply: Correct.
log\textsubscript{10} (\emph{p}-value) is depicted. The~dotted lines represent cut-off values for significantly changed genes (log\textsubscript{2} fold change $>$ 0.26 or $<$ $-$0.26, \emph{p}-value $<$ 0.05). A brown color depicts that a gene meets both cut-off criteria, a~blue color relates to meeting only the \emph{p}-value cut-off, an~orange color relates to meeting only the log$_{2}$ fold change cut-off, and a gray color indicates that a gene does not meet any of the criteria.}
\label{fig:fig1}
\end{figure}
\begin{paracol}{2}
%\linenumbers
\switchcolumn
\vspace{-12pt}

@@

\subsection{Toxicity-Related Pathways}
To gain more insight into the toxicity-related processes, the~overrepresented pathways were further categorized into pathways related to the apoptotic process, inflammatory response, DNA damage response, and/or oxidative stress. Often this kind of clustering of the pathways is performed manually. However, we implemented a gene-based approach. First, gene sets of four Gene~Ontology~(GO) terms were obtained, i.e.,~“apoptotic process”~(GO\hl{:}0006915, %MDPI: please confirm if space should be added after ``:''.
%Reply: Should not be added.
1269 genes), “inflammatory response”~(GO\hl{:}0006954, 569 genes),  “cellular response to DNA damage stimulus”~(GO\hl{:}0006974, 762 genes), and~“response to oxidative stress”~(GO\hl{:}0006979, 243 genes). Evidently, these processes are tightly connected. Whereas some can be causative for others, they are expected to overlap. Additional Figure~\ref{fig:figA1} shows the gene overlap between the four GO gene sets in a Venn diagram. Next, we calculated the gene overlap between the pathways from the WikiPathways pathway collection with the annotated genes of the four GO-terms. The~overlap was calculated by dividing the number of toxicity-related genes present in the pathway by the total number of genes present in the pathway. As~a cut-off we used 50$\%$ indicating that at least half of the pathway is directly linked to the GO-term of interest via their gene~overlap.

To select the desired gene overlap cut-off, we compared four cut-offs with each other and an overrepresentation analysis approach to see how many pathways these would give. The~cut-offs we used were 50\%, 60\%, 70\%, and 80\%. For~the overrepresentation analysis we looked for overrepresentation of the pathway genes in the gene lists of the GO-terms. The~results are shown in additional Table~\ref{tab:tabA1}. The~overrepresentation analysis \mbox{(\emph{p}-value~\hl{$<$}~0.05)} %Please confirm if this is correct.
%Reply: Was incorrect and it should be a single less than sign;
%       removed redundant <.
showed the most pathways to be linked to the four GO-terms; in~this case it showed 401, 250, 203, and 205 pathways overrepresented for the apoptotic process, inflammatory response, DNA damage, and oxidative stress GO-terms, respectively. The~80\% and 70\% gene overlap resulted in almost similar amounts of pathways compared to each other. They even resulted in 0 pathways for the inflammatory response GO-term. Moreover, it stands out that for the gene overlap approach only one pathway meets at least 50\% or higher gene overlap with the oxidative stress GO-term. Additionally, it is interesting to see that an overrepresentation analysis prompts hundreds of pathways compared to just dozens when looking at gene overlap. Comparing the results of the overrepresentation analysis and various cut-offs, the at least 50\% gene overlap cut-off was deemed best for our~approach.

After this filtering step we found that, out of the 1076 pathways in the human pathway collection, 66 pathways are related to the apoptotic process, 15 are linked to inflammatory response, 28 are linked to DNA damage, and 2 are linked to oxidative stress.
It is noticeable that there are only two oxidative stress pathways, i.e.,~Detoxification of Reactive Oxygen Species~(wikipathways:WP2824~\cite{WP2824})
%MDPI: added space, please confirm.
%Reply: Must not be added. Have a look at: https://doi.org/10.1038/sdata.2018.29 for reference.
and Oxidative Stress~(wikipathways:WP408~\cite{WP408}), that have at least 50$\%$ of their genes overlapping with the genes annotated to the GO-term ``response to oxidative stress''. Based on this finding we can argue that most pathways in the pathway database we have used describe different processes that result in oxidative stress or where oxidative stress is part of/has an influence on the process itself. However, these do not describe oxidative stress as a process.  Nevertheless, these pathways are of interest for our~analysis.

@@

\subsection{{Caco-2/HT29-MTX} Cells}
While for both concentrations the {Caco-2/HT29-MTX co-culture} shows the same number of overrepresented pathways, the~pathways compared between the low and high concentrations are not all the same. Of~the ten pathways, three are found for both concentrations, namely IL-5 Signaling pathway (wikipathways:WP127,~\cite{WP127}), IL-2 Signaling pathway (wikipathways:WP49,~\cite{WP49}) and Endometrial Cancer (wikipathways:WP4155,~\cite{WP4155}). 
Interestingly, for the low concentration the pathways related to DNA-damage and repair show up in the results. %check meaning has been retained. Added two words to retain the meaning of the sentence.
For~example, HDR Through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways:WP3567,~\cite{WP3567}), Nonhomologous End-Joining (NHEJ) (wikipathways:WP3550,~\cite{WP3550}), and DNA Double-Strand Break Response (wikipathways:WP3543,~\cite{WP3543}) are among these pathways. However, for the high concentration it is interesting to see that the Apoptosis pathway (wikipathways:WP254,~\cite{WP254}) shows up, which could be an indication of {Caco-2/HT29-MTX} cells possibly undergoing apoptotic processes upon exposure to TiO\textsubscript{2}-nanobelts. However, these results indicate that the studied processes in {Caco-2/HT29-MTX co-cultures} are not extensively affected by the administration of TiO\textsubscript{2}-nanobelts, for~either the low or the high concentration. This finding supports the conclusion based on the cell viability assay results in the original study, which showed that there was no significant decrease in cell viability for the {Caco-2/HT29-MTX co-culture} for both concentrations of TiO\textsubscript{2}-nanobelts~\cite{Tilton2013}.

\begin{figure}[H]
\includegraphics[width=1\linewidth]{fig2.pdf}
  \caption{Pathway-gene networks of altered toxicity-related pathways.
   The color of the nodes indicates to which GO-term the pathway is affiliated. Orange indicates apoptosis, blue indicates inflammation, pink indicates DNA-damage, and 
   red %check meaning has been retained. Correct.
   indicates oxidative stress. Gray nodes represent genes. Number (n) represents the number of significantly overrepresented pathways that are depicted in the networks.}
\label{fig:fig2}
\end{figure}
\unskip
\subsection{SAE Cells}
For the SAE cell line, the~low concentration only prompted three overrepresented pathways compared to 15 for the high concentration. Noticeable as well, only one of those three pathways was found to be overrepresented also for the high concentration, namely HDR through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways:WP3567,~\cite{WP3567}). This increase in pathways with a significantly increased overrepresentation of affected genes is an indication that upon exposure to the high concentration of TiO\textsubscript{2}-nanobelts, more biological processes related to toxicity are affected compared to the low~concentration.

Next, for~the SAE cell line exposed with the low concentration, we found three overrepresented pathways. As~for the high concentration we found 15 overrepresented pathways. This increase in pathways with a significantly increased overrepresentation of affected genes is an indication that more biological processes related to toxicity are affected compared to the low concentration upon exposure to the high concentration of TiO\textsubscript{2}-nanobelts . For~the low concentration the pathways HDR through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways:WP3567,~\cite{WP3567}), IL-6 Signaling pathway (wikipathways:WP364,~\cite{WP364}), and Regulation of TP53 Activity through Phosphorylation (wikipathways:WP3838,~\cite{WP3838}) are overrepresented. The~first pathway describes a process of DNA double-strand break repair, where the second pathway describes the signaling of the cytokine IL-6 and the last pathway describes the regulation of TP53. Except~for the first pathway, which is also found in the result for the high concentration, the~overrepresented pathways describe more general regulatory pathways rather than detailed process-describing pathways. However, for~the high concentration we see multiple pathways related to DNA damage, such as the aforementioned HDR through Homologous Recombination (HRR) or Single Strand Annealing (SSA) (wikipathways:WP3567,~\cite{WP3567}), DNA IR-damage and Cellular Response via ATR (wikipathways:WP4016,~\cite{WP4016}), DNA IR-Double Strand Breaks (DSBs) and Cellular Response via ATM (wikipathways:WP3959,~\cite{WP3959}), and DNA Mismatch Repair (wikipathways:WP531,~\cite{WP531}) among others. Moreover, the results prompted apoptosis-related pathways such as Apoptosis Modulation and Signaling (wikipathways:WP1772,~\cite{WP1772}) and Apoptosis Modulation by HSP70 (wikipathways:WP384,~\cite{WP384}). This is also an indication that the high concentration induced more distinct biological processes related to toxicity compared to the low concentration for the SAE cell~line.

\subsection{THP1 Cells}
For the THP1 cell line, there are only 10 significantly overrepresented pathways for the low concentration of TiO\textsubscript{2}-nanobelts. However, among~these pathways, the~pathways TNF Signaling (wikipathways:WP3380,~\cite{WP3380}), Apoptosis (wikipathways:WP254,~\cite{WP254}), Apoptotic Execution Phase (wikipathways:WP1784,~\cite{WP1784}), TNF alpha Signaling Pathway (wikipathways:WP231,~\cite{WP231}), and TNF-related Weak Inducer of Apoptosis (TWEAK) Signaling pathway (wikipathways:WP2036,~\cite{WP2036}) are present. This indicates that upon exposure to the low concentration of TiO\textsubscript{2}-nanobelts the THP1 cell line activates immune- and inflammation-related processes. This result can be explained as a general cell activity response since THP1 is a macrophage-like cell line. However, it can also be explained due to the effect of TiO\textsubscript{2}-nanobelts on the THP1 cell line in this case. TiO\textsubscript{2}-nanobelts are likely to induce toxic processes, even at a low concentration, which results in an inflammatory response of the THP1 cells. Similar inflammatory responses, like activation of Nf-$\kappa$B and production of TNF-$\alpha$, were seen in these cells upon exposure to other nanoparticles such as ZnO NM-110, SiO\textsubscript{2} NM-200, and Ag NM-300~\cite{Brzicova2019}.

Compared to the low concentration, the~high concentration yields more pathways with significant overrepresentation i.e.,~39 versus 10. Except~for the pathway Apoptotic Execution Phase (wikipathways:WP1784,~\cite{WP1784}) all other 9 pathways which showed up for the low concentration were found for the high concentration as well. In~addition to these pathways, pathways such as Oxidative Stress (wikipathways:WP408,~\cite{WP408}), Nanoparticle Triggered Regulated Necrosis (wikipathways:WP2513,~\cite{WP2513}), and Mismatch Repair (wikipathways:WP3381,~\cite{WP3381}) %please be consistent with capitalization of pathways throughout the paper (at present some have initial capitals and some do not). Changed to be consistent throughout manuscript.
and more were among the significant pathways. These results indicate that the THP1 cell line upon exposure to the high concentration of TiO\textsubscript{2}-nanobelts results in toxicity-related processes such as inflammation, DNA damage, and~oxidative stress. The~higher number of pathways that show up in the result together with the biological processes they depict indicate that the high concentration induces greater effects than the low concentration. This was also seen in the original paper where the low concentration induced a significant decrease in cell viability, while the high concentration induced an even greater decrease~\cite{Tilton2013}. Furthermore, it has also been shown that nanoparticles can activate similar processes, such as inflammatory processes and DNA damage, as~seen upon exposure to the high concentration~\cite{Brzicova2019,Huang2017}.

{The THP1 cell line is the cellular model that has the most inflammation-related pathways in the results: four for the low concentration and nine for the high concentration. All four of the low concentrations show up in the results of the high concentration as well. These pathways are Photodynamic-therapy-induced NF-kB Survival Signaling} (wikipathways:WP3617,~\cite{WP3617}), {Interleukin-10 Signaling} (wikipathways:WP4063,~\cite{WP4063}), {Fibrin Complement Receptor 3 Signaling Pathway} (wikipathways:WP4136,~\cite{WP4136}), {and Platelet-mediated Interactions With Vascular and Circulating Cells} (wikipathways:WP4462,~\cite{WP4462}).
{Nanoparticle-induced genotoxicity can arise and a distinction between primary and secondary genotoxicity can be made. Inflammation drives secondary toxicity as suggested by} Emma Åkerlund~et~al., {who showed that conditioned media from differentiated THP1 cells induced DNA-damage in HBEC after} 3~h~exposure~\cite{Akerlund2019}. {While there are inflammatory-related pathways in the results for the THP1 cell line, only the high concentration has three DNA-damage-related pathways in the results. The~presence of DNA-damage-related pathways and inflammatory-related pathways could indicate secondary genotoxicity for the THP1 cells upon exposure to the high concentration. However, more research is needed to address this exact mechanism.}

@@

To illustrate how the effects can be studied in more depth we visualized the log fold change of all six conditions on the Oxidative Stress pathway (wikipathways:WP408~\cite{WP408}), the~Apoptosis pathway (wikipathways:WP254~\cite{WP254}), and the DNA Mismatch Repair pathway (wikipathways:WP531~\cite{WP531}). This shows that for most genes in these pathways gene expression data are present (see additional Figures~\ref{fig:figA2} and~\ref{fig:figA3}).

@@

\begin{figure}[H]
\includegraphics[height=14cm,keepaspectratio]{fig3.png}
  \caption{\hl{Visualization} of log fold change on the Oxidative Stress pathway (wikipathways:WP408) for all six conditions.
{Cellular models} are depicted from left to right as {Caco-2/HT29-MTX}, SAE, and THP1. The~top row depicts the low concentrations whereas the bottom row depicts the high concentrations. {Color gradient for differential gene expression after exposure goes from blue (downregulated) via white (not changed) to red (upregulated).}}
\label{fig:fig3}
\end{figure}
\vspace{-12pt}

@@

\subsection{Reproducible Analysis Workflow}
The complete analysis is automated in R (version 3.6.3) and can easily be repeated with a new transcriptomics dataset or different selection focus. The~R scripts are available on GitHub (\url{https://github.com/laurent2207/TiO2-scripts} \hl{(accessed on 2021-08-28)}) and archived on Zenodo~\cite{TiO2-scripts}.
%Reply: date added

@@

\bibitem[Bohler \em{et~al.}(2020)Bohler, {Reactome Team}, Weitz, and
  Kutmon]{WP2824}
Bohler, A.; {Reactome Team}; Weitz, E.; Kutmon, M.
\newblock Detoxification of Reactive Oxygen Species (Homo Sapiens). 2020.
\newblock  Available online: \url{https://wikipathways.org/instance/WP2824_r117697} \hl{(accessed on 2021-08-28). } %Please add accessed date.
%Reply: date added
